Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-13 ms.tex
fix migration model; fix appendix D; fix ref to Altenberg2012; add Uri

% Title page
\title{Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention}

\author[a]{Yoav Ram}
\author[b]{Lee Altenberg}
\author[a]{Marcus W. Feldman}
\affil[a]{Department of Biology, Stanford University, Stanford, CA}
\affil[b]{Information and Computer Sciences, University of Hawai`i at M{\=a}noa, Honolulu, HI}

@@

The assumptions and conditions for the validity of the \emph{unified reduction principle} by neutral genetic modifiers are~\citep{Altenberg2017}:
(i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
A departure from the latter assumption occurs if two variation-producing processes interact~\citep{Feldman1980,Altenberg2012}.
Departures from the \emph{reduction principle} have also been demonstrated when conditions (i)-(iii) are not met, see for example~\citet{Holsinger1986} and references therein.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_i$ is $M_{i,k}$.
Therefore, the change in the frequencies of types $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}\label{eq:model}
\bar{w} f' = (I-C+MC)D f,
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{i,k} \ge 0$ for all $i,k$ and $\sum_{k=1}^n {M_{i,k}} = 1$ for all $j$; 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

% Migration model
\paragraph*{Migration model.}\label{sec:mig_model_1}

Here we assume a large population of haploids that occupies two demes.
Let the fitness of genotypes $A_k$ be $w_k$ in deme 1 and $q_k$ in deme 2 for $k=1,2,\ldots,n$.
The probability that an individual with genotype $A_k$ migrates between the deme is $m_k$; that is, migration rates depend on the genotypes.
Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, $\sum_{k=1}^{n}{x_k} = 1$, and $\sum_{k=1}^{n}{y_k} = 1$.

Following selection and migration the new frequencies $x_k'$ and $y_k'$ are given by
\begin{align}
x_k' = (1-m_k) \frac{w_k}{\bar{w}} x_k + m_k \frac{q_k}{\bar{q}} y_k \\
y_k' = m_k \frac{w_k}{\bar{w}} x_k + (1 - m_k) \frac{q_k}{\bar{q}} y_k
\end{align}
for $k=1,2,\ldots,n$, where $\bar{w} = \sum_{k \ge 0}{w_k x_k}$ and $\bar{q} = \sum_{k \ge 0}{q_k y_k}$ are the population mean fitnesses in deme 1 and 2, respectively.
This is an extension of a model presented by~\citet{Altenberg2017}.

To rephrase this model in terms of the general model, we define
\begin{equation}
\begin{aligned}
f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
C = diag[m_1, \ldots, m_n, m_1, \ldots, m_n], \\
D = diag[w_1, \ldots, w_n, q_1, \ldots, q_n], \\
M = \begin{bmatrix}
0 & I_n \\
I_n & 0
\end{bmatrix}
\end{aligned}
\end{equation}
where $I_n$ is the n-by-n identity matrix.
In this case, $M$ is reducible and so is $(I-C+MC)D$.

@@

Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}
Note that (i) the existence of $\bar{w}^*$ and $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix; 
(ii) the local stability of this equilibrium is proven in~\nameref{sec:AppD}; and
(iii) $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$. Then
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

Using \emph{Caswell's formula} (Eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
\end{equation}

@@

We have given several examples of variation-generating processes for which this principle applies -- namely mutation and migration -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type of the type or fitness of the transmitting individual, we expect that our results will hold.

@@

% Appendix B: Caswell's formula
\subsection*{Appendix B}\label{sec:AppB}

\citet{Caswell1978} gave a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
This is a useful formula~\citep{Caswell1978,Hermisson2002,Ram2012}, and therefore we reproduce it here.

\begin{lemma}[Caswell's formula]
Let $T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
Then the sensitivity of $\lambda$ to changes in any element $t$ of the matrix $T$ is
\begin{equation}\label{eq:Caswells_formula}
\frac{\partial \lambda}{\partial t} = 
u\tr \frac{\partial T}{\partial t} v
\end{equation}
\end{lemma}

@@

\emph{Fisher's reproductive values} can be understood as follows~\citep{Grafen2006}.
Consider the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and the corresponding frequencies are $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
The dynamics are
\begin{equation}
n(t) = (MD)^t n(0).
\end{equation}

@@

% Appendix D: inner stability
\subsection*{Appendix D}\label{sec:AppD}
\begin{lemma}[Stability of the equilibrium $v$]
The equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is locally stable.
\end{lemma}
\begin{proof}
Set $A=(I-C+MC)D$ and rewrite Eq.~\ref{eq:model} such that $f(t)$ is the frequency vector at time $t$ and
\begin{equation}
f(t) = \frac{Af(t-1)}{\norm{Af(t-1)}},
\end{equation}
where $\norm{f(t)} = \sum_{k=1}^{n}f_k(t)$.
Denote the eigenvalues of $A$ as $\lambda_1, \ldots, \lambda_n$ such that $\abs{\lambda_1} \ge \ldots \ge \abs{\lambda_n}$ and $\lambda_1$ is the \emph{Perron} leading eigenvalue, that is, the largest real eigenvalue with corresponding real non-negative right eigenvector $v$ ($\lambda_1$ is shown in~\nameref{sec:AppA} to be the stable population mean fitness $\bar{w}^*$).

First, note that the normalization $x \goesto \frac{1}{\norm{x}}$ can be applied once instead of $t$ times~\citep[Appendix~A]{Wakano2004}, that is,
\begin{equation}
f(t+k) = \frac{A^t f(k)}{\norm{A^t f(k)}}.
\label{eq:ftplus1}
\end{equation}

To check local stability, we make a perturbation $\epsilon$ to $v$, such that this perturbation is small enough so that when $v+\epsilon$ is written as a linear combination of $\{v^k\}_{k=1}^{n}$ the eigenvalues of $A$, the coefficient of $v=v^1$ is not zero.
That is, the frequency vector after the perturbation $f(0)=v+\epsilon$ can be written
\begin{equation}
f(0) = a_1 v^1 + \ldots + a_n v^n,
\label{eq:linear_combination}
\end{equation}
with $\sum_{k=1}^{n}{f_k(0)}=1$ and $a_1 \ne 0$.

We now check that $f(t) \goesto v$ as $t \goesto \infty$, which will indicate that $v$ is locally stable.
We write, using Eq.~\ref{eq:linear_combination},
\begin{multline}
A^t f(0) = 
a_1 A^t v^1 + a_2 A^t v^2 + \ldots + a_n A^t v^n = \\
a_1 \lambda_1^t v^1 + a_2 \lambda_2^t v^2 + \ldots + a_n \lambda_n^t v^n = \\
\lambda_1^t \bigg(a_ 1v^1 + a_2 \bigg(\frac{\lambda_2}{\lambda_1}\bigg)^t v^2 + \ldots +  a_n \bigg(\frac{\lambda_n}{\lambda_1}\bigg)^t v^n\bigg).
\end{multline}
Therefore,
\begin{equation}
\frac{A^t f(0)}{\lambda_1^t} \goesto a_1 v_1
\end{equation}
as $t \goesto \infty$, because $\abs{\frac{\lambda_k}{\lambda_1}} < 1$
for all $k=2, \ldots, n$.
Finally, 
\begin{equation}
f(t) = 
\frac{A^t f(0)}{\norm{A^t f(0)}} =
\frac{\frac{A^t f(0)}{\lambda_1^t}}{\norm{\frac{A^t f(0)}{\lambda_1^t}}} \goesto
\frac{a_1 v^1}{\norm{a_1 v^1}} = v,
\end{equation}
as $t \goesto \infty$, which completes the proof.
\end{proof}

>>>>

% Title page
\title{Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention}
% \title{A General Principle for the Evolution of Fitness-Associated Generation of Variation}

\author[a]{Yoav Ram}
\author[b]{Lee Altenberg}
\author[c]{Uri Liberman}
\author[a]{Marcus W. Feldman}
\affil[a]{Department of Biology, Stanford University, Stanford, CA}
\affil[b]{Information and Computer Sciences, University of Hawai`i at M{\=a}noa, Honolulu, HI}
\affil[c]{School of Mathematical Sciences, Tel Aviv University, Israel}

@@

The assumptions and conditions for the validity of the \emph{unified reduction principle} by neutral genetic modifiers are~\citep{Altenberg2017}:
(i) effectively infinite population size, (ii) constant-viability selection, (iii) a population at an equilibrium, and (iv) \emph{linear variation} -- the equal scaling of transition probabilities by the modifier.
A departure from the latter assumption occurs if two variation-producing processes interact~\citep{Feldman1980,Altenberg2012b}.
Departures from the \emph{reduction principle} have also been demonstrated when conditions (i)-(iii) are not met, see for example~\citet{Holsinger1986} and references therein.

@@

\paragraph*{General model.}\label{sec:general_model}
Consider a large haploid population with an arbitrary set of types $A_1, A_2, \ldots, A_n$.
The frequency and fitness of individuals of type $A_k$ are $f_k$ and $D_k$, respectively.
The probability that an individual of type $A_k$ will transition to some other type is $C_k$, and given a transition occurs, the probability that it will transition to type $A_j$ is $M_{j,k}$.
Therefore, the change in the frequencies of type $A_k$ is described by the transformation $f \to f'$: 
\begin{equation}
f'_k = (1-C_k)D_k f_k + \sum_{j=1}^{n}{C_j M_{k,j} D_j f_j},
\end{equation}
or in matrix form
\begin{equation}
\bar{w} f' = (I-C+MC)D f,
\label{eq:model}
\end{equation}
where $f=(f_1, f_2, \ldots, f_n)$ is a stochastic vector such that $\sum_{k=1}^n{f_k} = 1$;
$D$ is a diagonal matrix with diagonal elements $D_k$, where we can assume $0 < D_k \le 1$, i.e., fitness values are relative to the fitness of the fittest type; 
$C$ is a diagonal non-negative matrix with diagonal elements $C_k$;
$M$ is a column-stochastic matrix: $M_{j,k} \ge 0$ for all $j,k$ and $\sum_{j=1}^n {M_{j,k}} = 1$ for all $k$; 
$I$ is the identity matrix;
and $\bar{w}$ is the normalizing factor such that $\sum_{k=1}^n{f'_k}=1$ and is equal to the population mean fitness (see~\nameref{sec:AppA}).

@@

% Migration model
\paragraph*{Migration model.}\label{sec:mig_model_1}
% TODO explain why sum x y is 1 and wbar is sum xw yq
Here we assume a large haploid population of haploids that occupies two demes (the model can be easily extended to an arbitrary number demes).
Let the frequencies of $A_k$ be $x_k$ in deme 1 and $y_k$ in deme 2 for $k=1,2,\ldots,n$ with $x_k \ge 0$, $y_k \ge 0$, and $\sum_{k=1}^{n}{x_k + y_k} = 1$.

In this model, the fitness of the individual is determines by both its type and its deme, but the entire population comes together for reproduction, and therefore reproductive success is determined by competition between individuals of both demes.
Therefore, the fitness of genotypes $A_k$ is $w_{1,k}$ in deme 1 and $w_{2,k}$ in deme 2 for $k=1,2,\ldots,n$, and the population mean fitnesses of the entire population is $\bar{w} = \sum_{k \ge 0}{w_{1,k} x_k + w_{2,k} y_k}$.
In addition, we assume that the fitness of type $k$ is different in deme 1 and 2, that is, $w_{1,k} \ne w_{2,k}$ for any $k=1,\ldots,n$.

The probability that offspring of an individual from deme $a$ with type $A_k$ return to their parent home deme is $1-m_{j,k}$.
That is, migration rates also depend on both the type and the deme.

Following selection and migration the new frequencies $x_k'$ and $y_k'$ are given by
\begin{align}
x_k' = (1-m_{1,k}) \frac{w_{1,k}}{\bar{w}} x_k + m_{2,k} \frac{w_{2,k}}{\bar{w}} y_k \\
y_k' = m_{1,k} \frac{w_{1,k}}{\bar{w}} x_k + (1 - m_{2,k}) \frac{w_{2,k}}{\bar{w}} y_k
\end{align}
for $k=1,2,\ldots,n$.

To rephrase this model in terms of the general model, we define
\begin{equation}
\begin{aligned}
f = (x_1, \ldots, x_n, y_1, \ldots, y_n)\tr, \\
C = diag[m_{1,1}, \ldots, m_{1,n}, m_{2,1}, \ldots, m_{2,n}], \\
D = diag[w_{1,1}, \ldots, w_{1,n}, w_{2,1}, \ldots, w_{2,n}], \\
M = \begin{bmatrix}
0 & I_n \\
I_n & 0
\end{bmatrix}
\end{aligned}
\end{equation}
where $I_n$ is the n-by-n identity matrix.
In this case, $M$ is block matrix with each block an irreducible 2-by-2 matrix, and so is $(I-C+MC)D$.

@@

Write the equilibrium of the frequency vector $f$ in Eq.~\ref{eq:model} as $v$ and the stable population mean fitness as $\bar{w}^*$, then
\begin{equation}\label{eq:model_equilibrium}
\bar{w}^* v = (I-C+MC)D v.
\end{equation}
Note that (i) the existence of $\bar{w}^*$ and $v$ are guaranteed by the \emph{Perron-Frobenius theorem}~\cite[Appendix~A]{Otto2007} because $(I-C+MC)D$ is a non-negative matrix; 
(ii) the global stability of this equilibrium is proven in~\nameref{sec:AppD}; and
(iii) $\bar{w}^*$ is identified as the stable population mean fitness in~\nameref{sec:AppA}.

@@

\begin{lemma}[Mean fitness principle]\label{lemma:mfp}
Let $\bar{w}^*$ be the leading eigenvalue of $(I-C+MC)D$, $u$ and $v$ be the corresponding left and right eigenvectors, and $[M]_k$ the $k$-th column of $M$.
%TODO If $MCD$ is primitive, then
Then,
\begin{equation}\label{mfp}
\frac{\partial \bar{w}^*}{\partial C_k} = 
\frac{u_k v_k}{C_k} (\bar{w}^* - D_k),
\end{equation}
or in simpler terms,
\begin{equation}\label{mfp_sign}
sign\frac{\partial \bar{w}^*}{\partial C_k} = 
sign(\bar{w}^* - D_k).
\end{equation}

@@

Using the formula in~\citet{Caswell1978} (see eq.~\ref{eq:Caswells_formula} in~\nameref{sec:AppB}),
\begin{equation}
\frac{\partial \bar{w}^*}{\partial C_k} = 
u\tr \frac{\partial (I-C+MC)D}{\partial C_k} v.
\end{equation}

@@

% see Altenberg 2012 Ecological Monographs for a condition-dependent migration model
We have given several examples of variation-generating processes for which this principle applies -- namely mutation and migration -- but our model may apply to other processes as well.
For example, the \emph{reduction principle} applies to ecological models of dispersal, and~\citet{Gueijman2013} have demonstrated that even in homogeneous environments, fitness-associated dispersal increases the population mean fitness and is favored by selection over uniform dispersal.
% TODO see cultural transmission model in Lee's thesis
Similarly, if the transmission fidelity of culturally-transmitted traits depends on the type or fitness of the transmitting individual, we expect that our results will hold.

@@

% Appendix B: Caswell formula
\subsection*{Appendix B}\label{sec:AppB}

\citet{Caswell1978} gave a \emph{formula for the sensitivity of the population growth rate to changes in life history parameters}.
In this formula, the \emph{population growth rate} is the leading eigenvalue of the population transformation matrix $T$, the \emph{life history parameters} are elements of $T$, and the \emph{sensitivity} is the derivative of the former with respect to the latter.
This is a useful formula~\citep[ch.~10]{Caswell1978,Hermisson2002,Ram2012,Otto2007}, and therefore we reproduce it here.

\begin{lemma}
$T$ be a non-negative matrix with leading eigenvalue $\lambda$ and left and right eigenvectors $u$ and $v$ such that $\sum{v_k}=1$ and $u\tr v = \sum{u_k v_k} = 1$.
Then the sensitivity of $\lambda$ to changes in any element $t$ of the matrix $T$ is
\begin{equation}\label{eq:Caswells_formula}
\frac{\partial \lambda}{\partial t} = 
u\tr \frac{\partial T}{\partial t} v
\end{equation}
\end{lemma}

@@

\emph{Fisher's reproductive values} can be understood as follows~\citep[ch.~10]{Grafen2006,Otto2007}.
Consider the dynamics not of frequencies but of absolute population sizes such that the vector of the number of individuals of each type at time $t$ is $n(t)$ and the corresponding frequencies are $f_k(t) = n_k(t) / \sum_i{n_i(t)}$.
The dynamics are
\begin{equation}
n(t) = (MD)^t n(0).
\end{equation}

@@

% Appendix D: inner stability
\subsection*{Appendix D}\label{sec:AppD}
\begin{lemma}[Stability of the equilibrium $v$]
If $C$ and $D$ are positive diagonal matrices, and $M$ is block irreducible, then
the equilibrium $v$ of the system in Eq.~\ref{eq:model_equilibrium} is globally stable. \end{lemma}
\begin{proof}
Let $A=(I-C+MC)D$ and denote the leading eigenvalue of $A$ as $\bar{w}^*$.

According to the \emph{Perron-Frobenius theorem}, 
\begin{equation}
\bigg(\frac{A}{\bar{w}^*}\bigg)^t \goesto \frac{v u\tr}{u\tr v} = v u\tr,
\end{equation}
where $u$ and $v$ are the left and right Perron eigenvectors of $A$ such that 
\begin{equation}
\sum_{k=1}^{n}{v_k}=1, \quad
u\tr v=\sum_{k=1}^{n}{u_k v_k}=1,
\label{eq:sum_uv}
\end{equation}
and $v u\tr$ is the \emph{Perron projection} into the eigenspace corresponding to $\bar{w}^*$.

Therefore, for any positive frequency vector $f$
\begin{equation}
\bigg(\frac{A}{\bar{w}^*}\bigg)^t \cdot f \goesto v u\tr \cdot f = \alpha v,
\label{eq:perron_projection}
\end{equation}
where $\alpha \in \mathbb{R}$.
Using eqs.~\ref{eq:sum_uv} and~\ref{eq:perron_projection}
\begin{equation}
\alpha = \alpha u\tr v = u\tr \alpha v = u\tr v u\tr f = u\tr f = \sum_{k=1}^{n}{u_k f_k} > 0,
\label{eq:alpha0}
\end{equation}
because $f$ is positive and $u$ is non-negative.

Now, let $\norm{x} = \sum_{k=1}^{n}{x_k}$, and therefore $x_t \goesto x$ implies
\begin{equation}
\norm{x_t} \goesto \norm{x}
\label{eq:norm_limit}
\end{equation}
as $t \goesto \infty$.

Rewrite Eq.~\ref{eq:model} as 
\begin{equation}
f' = \frac{A f}{\norm{A f}} 
\label{eq:model_norm}
\end{equation}
Note that when applying eq.~\ref{eq:model_norm} $t$ times, the normalization can be applied once~\cite[Appendix~A]{Wakano2004}.

Therefore, for a positive frequency vector $f$, we can apply $f \goesto f'$ (Eq.~\ref{eq:model_norm}) $t$ times, and using eqs.~\ref{eq:perron_projection} and ~\ref{eq:norm_limit}, we have
\begin{multline}
\frac{A^t f}{\norm{A^t f}} = 
\frac{\frac{1}{(\bar{w}^*)^t} A^t f}{\frac{1}{(\bar{w}^*)^t} \norm{A^t f}} = 
\frac{\big(\frac{A}{\bar{w}^*}\big)^t \cdot f }{ \norm{ \big(\frac{A}{\bar{w}^*}\big)^t \cdot f }} \goesto
\frac{\alpha v}{\norm{\alpha v}} = v,
\end{multline}
as $t \goesto \infty$, as $\norm{v}=1$ (eq.~\ref{eq:sum_uv}) and $\alpha \ne 0$ (eq.~\ref{eq:alpha0}).
This completes the proof.
\end{proof}
